Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-09 manuscript/manuscript.tex
edits

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) in some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and normalize the entries for
heterozygosity (see Supplementary Materials Sections
\ref{supp:ind-depth-var-corr} and \ref{supp:matrix-correction}). These
covariances are expected to be zero when only drift is acting, as only
heritable variation for fitness can create covariance between allele frequency
changes in a closed population \parencite{Buffalo2019-io}.  Averaging across
the ten replicate temporal covariances matrices, we find temporal covariances
that are statistically significant (95\% block bootstraps CIs do not contain
zero), consistent with linked selection perturbing genome-wide allele frequency
changes over very short time periods. The covariances between all adjacent time
intervals are positive and then decay towards zero as we look at more distant
time intervals, as expected when directional selection affects linked variants'
frequency trajectories until ultimately linkage disequilibrium and the additive
genetic variance for fitness associated with neutral alleles decays
\parencite{Buffalo2019-io}. The temporal covariances per replicate are noisier
but this general pattern holds; see Supplementary Figure
\ref{suppfig:barghi-cov-panels}. \textcite{Barghi2019-qy}'s design means that
the covariances we see in adjacent time intervals are on average ten
generations apart, and given the temporal decay in covariance we see, the
covariances on shorter time-scales (e.g. if adjacent generations had been
sequenced) may well be higher yet (see Supplementary Material Section
\ref{supp:barghi-covs} for more details).

@@

\begin{align}
  G(t) = \frac{\sum_{i\ne j}^{t-1} \cov(\Delta p_i, \Delta p_j)}{\var(p_t - p_0)}
\end{align}
%
which estimates the effect of selection on allele frequency change between the
initial generation $0$ and some later generation $t$, which can be varied to
see how this quantity grows through time. Since \textcite{Barghi2019-qy}
experiment is sequenced every ten generations, in the numerator for the
covariance we use the allele frequency changes between adjacent timepoints,
which are ten generations apart. Consequently, this leads our measure $G(t)$ to
be strongly conservative, since the temporal covariances within each
ten-generation block are not directly observable, and thus are not included in
the numerator of $G(t)$. Still, we find a remarkably strong signal. Greater
than $20\%$ of total, genome-wide allele frequency change over 60 generations
is the result of selection (Figure \ref{fig:figure-1}B).

@@

We've calculated the convergent correlation for all rows of the replicate
covariance matrices. Like temporal covariances, we visualize these through time
(Figure \ref{fig:figure-2}A), with each line representing the convergent
correlation from a particular reference generation $s$ as it varies with $t$
(shown on the x-axis). In other words, each of the colored lines corresponds to
the like-colored row of the convergence correlation matrix (upper left in
Figure \ref{fig:figure-2}A). We find these convergent covariances are
relatively weak, and decay very quickly from an initial value of about 0.1
(95\% block bootstrap confidence intervals $[0.094, 0.11]$) to around 0.01
(95\% CIs [0.0087, 0.015]) within 20 generations. This suggests that while
a reasonable fraction of the initial response is shared over the replicates,
this is followed by a rapid decay, a result consistent with the primary
finding of the original \textcite{Barghi2019-qy} study: that alternative loci
contribute to longer term adaptation across the different replicates.

@@

\clearpage
\subsubsection{\textcite{Barghi2019-qy} Empirical Null and Windowed Covariance Distributions}

  \label{suppfig:barghi-tailprobs-seqid}
\end{figure}

@@

\end{document}

>>>>

We first analyzed \textcite{Barghi2019-qy}, an evolve-and-resequence study with
ten replicate populations exposed to a high temperature lab environment and
evolved for 60 generations, and sequenced every ten generations. Using the
seven timepoints and ten replicate populations, we estimated the genome-wide $6
\times 6$ temporal covariance matrix $\mathbf{Q}$ for each of the ten
replicates. Each row of these matrices represent the temporal covariance
$\cov(\Delta_{_{10}} p_s, \Delta_{_{10}} p_t)$, between the allele frequency
change (in ten-generation intervals, denoted $\Delta_{_{10}} p_t$) in some
initial reference generation $s$ (the row of the matrix), and some later
timepoint $t$ (the column of the matrix). We corrected these matrices for
biases created due to sampling noise, and normalize the entries for
heterozygosity (see Supplementary Materials Sections
\ref{supp:ind-depth-var-corr} and \ref{supp:matrix-correction}). These
covariances are expected to be zero when only drift is acting, as only
heritable variation for fitness can create covariance between allele frequency
changes in a closed population \parencite{Buffalo2019-io}.  Averaging across
the ten replicate temporal covariances matrices, we find temporal covariances
that are statistically significant (95\% block bootstraps CIs do not contain
zero), consistent with linked selection perturbing genome-wide allele frequency
changes over very short time periods. The covariances between all adjacent time
intervals are positive and then decay towards zero as we look at more distant
time intervals (Figure \ref{fig:figure-1} A), as expected when directional
selection affects linked variants' frequency trajectories until ultimately
linkage disequilibrium and the additive genetic variance for fitness associated
with neutral alleles decays \parencite{Buffalo2019-io}. The temporal
covariances per replicate are noisier but this general pattern holds; see
Supplementary Figure \ref{suppfig:barghi-cov-panels}.
\textcite{Barghi2019-qy}'s design means that the covariances we see in adjacent
time intervals are on average ten generations apart, and given the temporal
decay in covariance we see, the covariances on shorter time-scales (e.g. if
adjacent generations had been sequenced) may well be higher yet (see
Supplementary Material Section \ref{supp:barghi-covs} for more details).

@@

\begin{align}
  G(t) = \frac{\sum_{i\ne j}^{t-1} \cov(\Delta p_i, \Delta p_j)}{\var(p_t - p_0)}
\end{align}
%
which estimates the effect of selection on allele frequency change between the
initial generation $0$ and some later generation $t$, which can be varied to
see how this quantity grows through time. Since \textcite{Barghi2019-qy}
experiment is sequenced every ten generations, in the numerator for the
covariance we use the allele frequency changes between adjacent timepoints,
which are ten generations apart. Consequently, this leads our measure $G(t)$ to
be strongly conservative, since the temporal covariances within each
ten-generation block are not directly observable, and thus are not included in
the numerator of $G(t)$. Still, we find a remarkably strong signal. Greater
than $20\%$ of total, genome-wide allele frequency change over 60 generations
is the result of selection (Figure \ref{fig:figure-1} B).

@@

We've calculated the convergent correlation for all rows of the replicate
covariance matrices. Like temporal covariances, we visualize these through time
(Figure \ref{fig:figure-2} A), with each line representing the convergent
correlation from a particular reference generation $s$ as it varies with $t$
(shown on the x-axis). In other words, each of the colored lines corresponds to
the like-colored row of the convergence correlation matrix (upper left in
Figure \ref{fig:figure-2} A). We find these convergent covariances are
relatively weak, and decay very quickly from an initial value of about 0.1
(95\% block bootstrap confidence intervals $[0.094, 0.11]$) to around 0.01
(95\% CIs [0.0087, 0.015]) within 20 generations. This suggests that while
a reasonable fraction of the initial response is shared over the replicates,
this is followed by a rapid decay, a result consistent with the primary
finding of the original \textcite{Barghi2019-qy} study: that alternative loci
contribute to longer term adaptation across the different replicates.

@@

@@

\subsubsection{\textcite{Barghi2019-qy} Empirical Null and Windowed Covariance Distributions}

  \label{suppfig:barghi-tailprobs-seqid}
\end{figure}
\end{document}
